---
title: "Why cograph?"
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE, 
  warning = FALSE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.dpi = 300,
  dpi = 300
)
```


```{r setup}
library(tna)
library(cograph)
```
 

## Do we need another network package?

R already has igraph for computation and qgraph for psychometric networks. tidygraph tries to unify things with a dplyr grammar. So why write another one?

The short answer: none of these packages let you go from a raw matrix to a filtered, annotated, publication-ready figure without switching between ecosystems, reformatting objects, or writing boilerplate. cograph tries to close that gap.

This document walks through the specific things cograph does differently, with working examples.

## Filtering and selecting — like data frames

In igraph, subsetting a network means calling `induced_subgraph()`, `delete_edges()`, or indexing into `V(g)` and `E(g)`. In tidygraph, you need `activate(nodes) |> filter(...)`. Both approaches require you to think about the object structure.

cograph lets you filter nodes and edges with expressions that look like `subset()`:

```{r}
mat <- matrix(c(
  0.0, 0.5, 0.8, 0.1, 0.0,
  0.3, 0.0, 0.2, 0.6, 0.4,
  0.7, 0.1, 0.0, 0.3, 0.5,
  0.0, 0.4, 0.2, 0.0, 0.9,
  0.1, 0.3, 0.6, 0.8, 0.0
), 5, 5, byrow = TRUE)
rownames(mat) <- colnames(mat) <- c("Read", "Write", "Plan", "Code", "Test")
net <- as_cograph(mat)
```

Keep only edges above a threshold:

```{r}
strong <- filter_edges(net, weight > 0.5)
get_edges(strong)
```

Keep nodes that have high degree *and* high PageRank — the centrality measures are computed on the fly:

```{r}
hubs <- filter_nodes(net, degree >= 3 & pagerank > 0.15)
get_nodes(hubs)
```

### Structural selections

`select_nodes()` goes further. It knows about components, neighborhoods, articulation points, and k-cores — and it computes them lazily (only what your expression actually references):

```{r}
# Top 3 nodes by betweenness
top3 <- select_top(net, n = 3, by = "betweenness")
get_nodes(top3)
```

```{r}
# Ego network: everything within 1 hop of "Code"
ego <- select_neighbors(net, of = "Code", order = 1)
get_nodes(ego)
```

For edges, you can select by structure too:

```{r}
# Edges involving "Code"
code_edges <- select_edges_involving(net, nodes = "Code")
get_edges(code_edges)
```

```{r}
# Top 5 edges by weight
top5 <- select_top_edges(net, n = 5)
get_edges(top5)
```

```{r}
# Edges between two node sets
between <- select_edges_between(net,
  set1 = c("Read", "Write"),
  set2 = c("Code", "Test")
)
get_edges(between)
```

### Format in, same format out

If you pass a matrix, you can get a matrix back:

```{r}
filter_edges(mat, weight > 0.5, keep_format = TRUE)
```

Same for igraph objects. No lock-in.

## Centrality — one call, tidy table

igraph requires a separate function for each centrality measure, and each returns a different format. `page_rank(g)` gives you a list with `$vector`. `betweenness(g)` gives a named numeric. Building a comparison table takes 10+ lines.

In cograph, one call with no arguments returns all 34 measures as a tidy data frame:

```{r}
centrality(net, digits = 3)
```

That includes degree, strength, betweenness, closeness, PageRank, eigenvector, but also less common ones like load, current-flow betweenness, voterank, percolation, diffusion, and leverage — all computed natively, without extra packages.

If you only need a subset:

```{r}
centrality(net, measures = c("degree", "betweenness", "pagerank"), digits = 3)
```

Need just one measure as a named vector? Use the wrapper:

```{r}
centrality_pagerank(net)
```

```{r}
centrality_betweenness(net)
```

You can normalize, sort, and round in the same call:

```{r}
centrality(net, measures = c("degree", "betweenness", "pagerank"),
           normalized = TRUE, sort_by = "pagerank", digits = 3)
```

### Edge-level centrality

```{r}
edge_centrality(net, sort_by = "betweenness", digits = 3)
```

### Network-level summary

One-row data frame with density, diameter, transitivity, centralization, reciprocity, and more:

```{r}
network_summary(net, digits = 3)
```

Add `detailed = TRUE` for mean/sd of node-level measures, or `extended = TRUE` for girth, radius, global efficiency, etc.

## Community detection — one call

igraph has `cluster_louvain()`, `cluster_walktrap()`, etc. — different function per algorithm, inconsistent parameter names. cograph wraps all of them behind one function with a default:

```{r}
# Undirected network for community detection
sym <- (mat + t(mat)) / 2
diag(sym) <- 0
cograph::communities(sym)
```

Pick a different algorithm by name, or use two-letter shorthands:

```{r}
cograph::communities(sym, method = "walktrap")
com_fg(sym)   # fast greedy
com_im(mat)   # infomap (works on directed too)
```

If you just want a node-to-community data frame:

```{r}
detect_communities(sym, method = "walktrap")
```

### Quality and significance

How good is the partition?

```{r}
comm <- com_wt(mat)
det <- detect_communities(mat, method = "walktrap")
cluster_list <- split(det$node, det$community)
cqual(mat, cluster_list)
```

Is the modularity significantly higher than chance? Permutation test against a null model:

```{r}
csig(mat, comm, n_random = 200, seed = 1)
```

### Compare two solutions

```{r}
comm2 <- com_fg(mat)
compare_communities(comm, comm2, method = "nmi")
```

### Consensus clustering

Run a stochastic algorithm many times and threshold the co-occurrence matrix:

```{r}
com_consensus(mat, method = "infomap", n_runs = 50, seed = 1)
```

## Format interoperability

cograph accepts matrices, edge lists, igraph, statnet, qgraph, and tna objects natively. And it converts back:

```{r}
net <- as_cograph(mat)

# To igraph
g <- to_igraph(net)
class(g)

# To matrix
m <- to_matrix(net)
m[1:3, 1:3]

# To edge list
head(to_df(net))
```

No format lock-in. Use cograph for what it's good at, convert back when you need something else.

## Robustness, motifs, backbone

A few more things that would otherwise require separate packages:

### Robustness analysis

How does the network hold up when you remove nodes by betweenness (targeted attack) vs. random failure?

```{r}
rob <- robustness(mat, measure = "betweenness", strategy = "sequential", seed = 1)
rob_rand <- robustness(mat, measure = "random", n_iter = 50, seed = 1)

# Area under curve — higher means more robust
robustness_auc(rob)
robustness_auc(rob_rand)
```

### Disparity filter (backbone extraction)

Keep only edges that carry a disproportionate share of a node's weight:

```{r}
backbone <- disparity_filter(mat, level = 0.5)
backbone
```

### Motif census

Count triad types with significance testing against a configuration model:

```{r}
motif_census(mat, n_random = 100)
```

## Working with probabilistic networks

cograph was built with transition networks in mind — matrices where rows sum to 1 and edges are probabilities, not just weights. This matters in a few places.

### Smart weight inversion

Path-based centrality measures (betweenness, closeness, harmonic) need distances, not probabilities. For transition networks, a high-probability edge should mean *short* distance. cograph handles this automatically:

- For tna objects: weights are inverted (distance = 1/probability)
- For other networks: weights are used as-is (standard igraph convention)

No manual toggle needed.

### First-class tna support

If you use the [tna](https://cran.r-project.org/package=tna) package for Transition Network Analysis, cograph understands its objects directly:

```{r}
model <- tna(group_regulation)
splot(model)
```

Bootstrap results render automatically — significant transitions as solid edges, non-significant as dashed:

```{r}
boot <- bootstrap(model, iter = 1000)
splot(boot)
```

Permutation test results get color-coded by group effect:

```{r}
model1 <- tna(group_regulation[1:1000,])
model2 <- tna(group_regulation[1001:2000,])

perm <- permutation_test(model1, model2, iter = 1000)
splot(perm)
```

Group comparisons with `plot_compare()` show element-wise probability differences, with donuts on nodes for initial state shifts.

### qgraph compatibility

Researchers coming from qgraph can use familiar parameter names — `vsize`, `asize`, `edge.color`, etc. — they are translated automatically. When both the cograph name and the qgraph alias are present, the cograph name wins.

### Nestimate integration

cograph also plots [Nestimate](https://cran.r-project.org/web/packages/Nestimate/index.html) objects (bootstrap forests, permutation results, glasso networks) without importing the package — dispatch is by class name only.

## Summary

cograph is not trying to replace igraph's graph algorithms or tidygraph's data manipulation. It fills a different gap: going from data to a filtered, annotated, publication-ready network figure with minimal code, while staying interoperable with everything else in the R network ecosystem.

The main ideas:

- **Filter and select** with data-frame-like expressions, not object-specific APIs
- **Centrality** as a tidy data frame, not a list of separate calls
- **Community detection** behind one function with shorthands, quality metrics, and significance tests
- **Statistical annotation** (CIs, p-values, stars) directly on the figure
- **No format lock-in** — matrix/igraph/statnet/tna in and out
- **Probabilistic networks** handled correctly out of the box
