---
title: "Plotting TNA Models with splot"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.dpi = 300,
  dpi = 300
)
library(cograph)
library(tna)
```

# 1. Single TNA Model — group_regulation

Build a TNA model from the `group_regulation` dataset (2000 sequences, 9 states).

```{r tna-model, results = F}
data(group_regulation)
model <- tna(group_regulation)
```

## Basic Network Plot

```{r plot-basic, fig.width=9, fig.height=8}
splot(model,
      title = "Group Regulation TNA",
      minimum = 0.05)
```

# 2. Bootstrap Analysis

Run bootstrap resampling (1000 iterations) to assess edge significance and confidence intervals.

```{r bootstrap, cache=TRUE}
set.seed(42)
boot <- bootstrap(model, iter = 1000)
```

```{r boot-summary}
sig_edges <- boot$summary[boot$summary$sig, ]
cat(sprintf("Significant edges: %d / %d\n", nrow(sig_edges), nrow(boot$summary)))
head(sig_edges[order(sig_edges$p_value), ], 10)
```

## Bootstrap — Significant Edges Only

```{r plot-boot-sig, fig.width=9, fig.height=8}
splot(boot,
      display = "significant",
      title = "Bootstrap — Significant Edges",
      show_stars = TRUE)
```

## Bootstrap — Full Network with Styling

Non-significant edges shown as dashed gray lines.

```{r plot-boot-styled, fig.width=9, fig.height=8}
splot(boot,
      display = "styled",
      title = "Bootstrap — Styled (sig=solid, nonsig=dashed)",
      show_stars = TRUE,
      threshold  = 0.02)

```

## Bootstrap — Confidence Intervals

```{r plot-boot-ci, fig.width=9, fig.height=8}
splot(boot,
      display = "ci",
      title = "Bootstrap — With Confidence Intervals",
      show_ci = TRUE,
      minimum = 0.05)
```
```{r tna-grouped, fig.width=5, fig.asp=1,out.width="50%",fig.pos="center",fig.align = 'center' }
plot_bootstrap_forest(boot, layout = "grouped",
  title = "Human-AI Interaction — Grouped by Source Node")
```

# 3. Simulated Group TNA + Permutation Test

Generate two group TNA networks then compare with permutation testing.

```{r group-tna}
set.seed(123)
group_models <- group_tna(group_regulation, 
                          group = c(rep("H", 1000), rep("L", 1000)))
```

## Plot Each Group

```{r plot-groups, fig.width=14, fig.height=7}
par(mfrow = c(1, 2))
splot(group_models[[1]], title = "Group 1", minimum = 0.05)
splot(group_models[[2]], title = "Group 2", minimum = 0.05)
```

## Difference Network

```{r plot-compare, fig.width=9, fig.height=8}
cograph::plot_compare(
  group_models[[1]], group_models[[2]],
  title = "Group 1 vs Group 2 — Difference Network")
```

## Permutation Test

```{r permutation, cache=TRUE}
set.seed(42)
perm <- tna::permutation_test(group_models[[1]], group_models[[2]], iter = 1000)
```

```{r perm-summary}
perm
```

## Permutation Test — Network

```{r plot-perm, fig.width=9, fig.height=8}
cograph::plot_permutation(perm,
      title = "Permutation Test — Significant Differences",
      show_nonsig = TRUE)
```
