---
title: "Bootstrap Forest Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstrap Forest Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 7,
  message = FALSE, warning = FALSE
)
library(cograph)
library(Nestimate)
```

## Overview

`plot_bootstrap_forest()` visualises bootstrapped edge weights and confidence
intervals for any network estimated with `bootstrap_network()` or
`boot_glasso()`.  Three layouts are available:

| Layout | Best for |
|--------|----------|
| `"linear"` | Many edges, precise comparison |
| `"circular"` | Medium networks, publication figures |
| `"grouped"` | Source-node grouping, colour by community |

`plot_edge_diff_forest()` visualises **pairwise edge weight differences** from
a `boot_glasso` object. Four layouts: `"linear"`, `"circular"`, `"chord"`,
`"tile"`.

---

## 1. TNA Network (relative transitions)

```{r tna-boot}
net_tna  <- build_network(human_wide, method = "relative")
boot_tna <- bootstrap_network(net_tna, iter = 200, seed = 42)
```

### Linear
```{r tna-linear, fig.height=10}
plot_bootstrap_forest(boot_tna,
  title    = "Human-AI Interaction Network",
  subtitle = "95% bootstrap CI  |  200 iterations")
```

### Circular
```{r tna-circular, fig.asp=1}
plot_bootstrap_forest(boot_tna, layout = "circular",
  title = "Human-AI Interaction Network — Circular")
```

### Grouped Radial
```{r tna-grouped, fig.width=10, fig.asp=1}
plot_bootstrap_forest(boot_tna, layout = "grouped",
  title = "Human-AI Interaction — Grouped by Source Node")
```

---

## 2. Glasso Network (partial correlations)

```{r srl-boot}
net_srl  <- build_network(srl_strategies, method = "glasso")
boot_srl <- boot_glasso(net_srl, iter = 200, seed = 42)
```

### Linear
```{r srl-linear, fig.height=10}
plot_bootstrap_forest(boot_srl,
  title = "SRL Strategies — Partial Correlation Network")
```

---

## 3. Edge Difference Plots (glasso)

Compare whether pairs of edges have significantly different weights.

### Tile Heatmap
```{r tile, fig.asp=1}
plot_edge_diff_forest(boot_srl, layout = "tile",
  title = "Edge Differences — Tile")
```

### Linear Forest
```{r ediff-linear, fig.height=10}
plot_edge_diff_forest(boot_srl, layout = "linear", n_top = 25,
  title = "Edge Differences — Linear")
```

### Chord Diagram
```{r chord, fig.width=10, fig.asp=1}
plot_edge_diff_forest(boot_srl,
  layout       = "chord",
  nonzero_only = TRUE,
  show_nonsig  = TRUE,
  title        = "Edge Differences — Chord",
  subtitle     = "Node colour = degree  |  ribbon = strength of difference")
```

---

## 4. Grouped Networks

Compare bootstrap CIs across groups in one plot.

```{r group-boot}
nets_grp  <- build_network(group_regulation_long,
  method = "relative", actor = "Actor",
  action = "Action",  time  = "Time",
  group  = "Achiever")
boots_grp <- bootstrap_network(nets_grp, iter = 200, seed = 42)
```

```{r group-forest, fig.height=10}
plot_bootstrap_forest(boots_grp,
  title    = "Group Regulation — High vs Low Achievers",
  subtitle = "95% bootstrap CI  |  200 iterations per group")
```
