---
title: "Network Estimation and Visualization with Nestimate + cograph"
description: |
    From raw sequence data to publication-ready statistical network plots ---
    build, bootstrap, compare, and visualize using Nestimate and cograph together.
date: "2026-04-01"
categories:
  - tutorial
  - R
  - visualization
  - Nestimate
warning: false
message: false
fig.align: center
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 8
    fig.height: 7
    dpi: 150
    comment: ""
---

```{r}
#| label: setup
#| include: false
library(Nestimate)
library(tna)
library(cograph)
```

# Introduction

**Nestimate** is a general-purpose network estimation package. It provides a unified `build_network()` interface that accepts raw sequence data and constructs networks using multiple methods: transition probabilities, partial correlations, regularized networks (GLASSO), co-occurrence, and more. The result is a `netobject` that works directly with cograph's `splot()` — no conversion needed.

Beyond estimation, Nestimate provides bootstrap stability analysis, permutation-based group comparisons, higher-order network construction (HON, HYPA), and motif analysis. All output objects dispatch through `splot()` automatically.


# Data

We use `group_regulation_long` — a long-format dataset of 9 collaborative regulation behaviors coded across student groups, with achievement group labels (High vs Low).

```{r}
#| label: data
data("group_regulation_long")
head(group_regulation_long)
```


# Estimating Networks

`build_network()` estimates a network from raw data. The `method` argument selects the estimator.

```{r}
#| label: build-network
net <- build_network(group_regulation_long,
  action = "Action", actor = "Actor", method = "relative")
net
```

`splot()` renders the `netobject` directly:

```{r}
#| label: fig-base
#| fig-cap: "Collaborative regulation — relative transition probabilities"
splot(net)
```

One line builds different network types:

```{r}
#| label: estimators
#| eval: false
build_network(group_regulation_long, action = "Action", actor = "Actor", method = "frequency")
build_network(group_regulation_long, action = "Action", actor = "Actor", method = "attention")
build_network(group_regulation_long, action = "Action", actor = "Actor", method = "co_occurrence")
build_network(group_regulation_long, action = "Action", actor = "Actor", method = "glasso")
build_network(group_regulation, method = "pcor")
```


# Bootstrap Stability

`bootstrap_network()` resamples the data and refits the model at each iteration, identifying which edges are stable and which are sampling artifacts.

```{r}
#| label: run-bootstrap
#| cache: true
set.seed(265)
boot <- bootstrap_network(net, iter = 1000)
boot
```

## Network plots

Significant edges appear solid; non-significant edges appear dashed grey.

```{r}
#| label: fig-boot-styled
#| fig-cap: "Solid = stable edges, dashed = unstable"
splot(boot)
```

```{r}
#| label: fig-boot-sig
#| fig-cap: "Significant edges only"
splot(boot, display = "significant")
```

```{r}
#| label: fig-boot-stars
#| fig-cap: "With significance stars"
splot(boot, show_stars = TRUE)
```

```{r}
#| label: fig-boot-ci
#| fig-cap: "With 95% confidence intervals"
splot(boot, show_ci = TRUE)
```

## Forest plots

Forest plots show edge weights and confidence intervals in a compact format. Three layouts are available:

```{r}
#| label: fig-forest-linear
#| fig-cap: "Linear forest plot"
#| fig-height: 11
#| fig-width: 6
plot_bootstrap_forest(boot)
```

```{r}
#| label: fig-forest-circular
#| fig-cap: "Circular forest plot"
plot_bootstrap_forest(boot, layout = "circular")
```

```{r}
#| label: fig-forest-grouped
#| fig-cap: "Grouped by source node"
plot_bootstrap_forest(boot, layout = "grouped")
```


# Group Networks

Pass `group =` to `build_network()` to estimate a separate network per group. `splot()` renders all groups in an automatic grid.

```{r}
#| label: build-group
grp_net <- build_network(group_regulation_long, method = "relative",
  action = "Action", actor = "Actor", time = "Time", group = "Achiever")
names(grp_net)
```

```{r}
#| label: fig-group-default
#| fig-cap: "High vs Low achievers on a common weight scale"
#| fig-width: 12
#| fig-height: 6
splot(grp_net)
```

`common_scale = TRUE` (default) makes edge widths directly comparable across panels:

```{r}
#| label: fig-group-noscale
#| fig-cap: "Each group on its own scale"
#| fig-width: 12
#| fig-height: 6
splot(grp_net, common_scale = FALSE)
```


# Permutation Testing

`permutation_test()` tests whether each edge differs significantly between two groups.

```{r}
#| label: run-perm
#| cache: true
set.seed(265)
perm <- Nestimate::permutation_test(grp_net$High, grp_net$Low, iter = 1000)
perm
```

```{r}
#| label: fig-perm-sig
#| fig-cap: "Green = High > Low, red = Low > High (p < 0.05)"
splot(perm)
```

```{r}
#| label: fig-perm-nonsig
#| fig-cap: "All edges — significant solid, non-significant dashed"
splot(perm, show_nonsig = TRUE)
```

```{r}
#| label: fig-perm-effect
#| fig-cap: "With significance stars and effect sizes"
splot(perm, show_stars = TRUE, show_effect = TRUE)
```


# Motifs

Nestimate objects work directly with cograph's motif analysis. `motifs()` counts MAN triad types; `subgraphs()` identifies specific node triples.

```{r}
#| label: motifs-census
mot <- cograph::motifs(net, significance = TRUE, n_perm = 100, seed = 42)
mot
```

```{r,fig.width=4, fig.height=2}
#| label: fig-motifs-types
#| fig-cap: "Triad type distribution"
plot(mot, type = "types")
```

```{r,fig.width=4, fig.height=2}
#| label: fig-motifs-sig
#| fig-cap: "Over- and under-represented triad types"
plot(mot, type = "significance")
```

Named subgraphs — which states form which patterns:

```{r}
#| label: subgraphs
sg <- cograph::subgraphs(net, significance = FALSE, top = 10)
sg
```

```{r}
#| label: fig-motifs-patterns
#| fig-cap: "MAN triad pattern diagrams"
#| fig-width: 10
#| fig-height: 6
plot(mot, type = "patterns")
```


# Higher-Order Networks

When first-order Markov models are insufficient, Nestimate detects higher-order sequential dependencies.

## HON

`build_hon()` identifies transitions where prior context changes the outgoing distribution:

```{r}
#| label: hon
hon <- build_hon(net)
hon
```

## HYPA

`build_hypa()` detects paths that occur significantly more or less often than expected:

```{r}
#| label: hypa
hypa <- build_hypa(net)
table(hypa$scores$anomaly)
```

```{r}
#| label: hypa-over
over <- hypa$scores[hypa$scores$anomaly == "over", ]
head(over[order(-over$ratio), c("path", "ratio", "observed", "expected")], 8)
```

## Simplicial visualization

```{r}
#| label: fig-simplicial
#| fig-cap: "Top higher-order pathways"
#| fig-width: 10
#| fig-height: 10
plot_simplicial(net, max_pathways = 6, title = "Higher-Order Pathways")
```

```{r}
#| label: fig-simplicial-dismantled
#| fig-cap: "Individual pathway panels"
#| fig-width: 12
#| fig-height: 12
plot_simplicial(net, max_pathways = 9, dismantled = TRUE, ncol = 3)
```


# The Dispatch Chain

Every Nestimate object routes through `splot()`:

```{r}
#| label: dispatch
#| eval: false
splot(net)      # netobject       → network plot
splot(boot)     # net_bootstrap   → stability-styled
splot(grp_net)  # netobject_group → multi-panel grid
splot(perm)     # net_permutation → difference network
```

All standard `splot()` arguments (`layout`, `node_fill`, `edge_color`, `title`, etc.) work with every object type.


# References {.unnumbered}

- Saqr, M., López-Pernas, S., Törmänen, T., Kaliisa, R., Misiejuk, K., & Tikka, S. (2025). Transition Network Analysis: A Novel Framework for Modeling, Visualizing, and Identifying the Temporal Patterns of Learners and Learning Processes. In *Proceedings of the 15th International Learning Analytics and Knowledge Conference (LAK '25)* (pp. 351--361). ACM. <https://doi.org/10.1145/3706468.3706513>

- Tikka, S., López-Pernas, S., & Saqr, M. (2025). tna: An R Package for Transition Network Analysis. *Applied Psychological Measurement*. <https://doi.org/10.1177/01466216251348840>

- cograph: <https://github.com/sonsoleslp/cograph>
- Nestimate: <https://github.com/mohsaqr/Nestimate>
